---
title: "Annotation 5"
output: html_document
date: "2026-06-01"
---

# Tidyverse

Tidyverse is a collection of packeges for R to better organize and work eficcietnt through projects. Thr core packages are the following ones:

* readr: data import
* tidyr: data tidying
* tibble: modern re-imagining of data frames
* dplyr: data manipulation
* stringr: for strings
* ggplot2: data visualisation
* purrr: functional programming
* forcats: dealing factors

There are official cheat-sheets for each package.

read_csv() for reading the comma-seperated values
if parsing, the best option might be to parse columns after loading the data to expolore the data first

A tibble is a like a dataframe, but with more enforcements and informations (e.g. dimensions and variable parameter) -> keeping the data consistent and ensure compatibility and efficiency in a project.
Compared to a dataframe (data.frame()), it lists the dimensions of the table and specifys the datatypes of each column
To convert a dataframe into a tibble, the command as_tibble(df) will do it

## Reshape data:
The command gather() takes multiple columns and collapses them into two new columns: a Key (the original column names) and a Value (the data inside those columns). It converts from wide to long format and is the standard requirement for  ggplot2 and performing statistical modeling
The command spread() does the exact opposite: it takes values from a "Long" list and distributes them into separate, side-by-side columns. In converts from long to wide, each category gets its own column. This version is often easier to read and preferred for summarizing results 
expand() makes a dataframe with every possible combination of the chosen variables
gather() and spread() are the older versions of this command. The new versions are pivot_longer() and pivot_wider()

**Dealing with NA's:** 
drop_na(table) -> deletes all rows with NA (may delete lots of data)
fill(table, columnName) -> fills the NA's whti the most recent value (value above)
replace_na(table, replace = list(columnName = value)); often used inside mutate()

**Unite and seperate:**
separate(data, col, into =c(colName1, colName2), sep = "/") is the argument to split one cell into multiple cells. The agrument for the separator needs to be specified (in this case with "/").
unite(data, colName1, colName2, col = newColumn, sep = ":") merges multiple cells into one. The intended separator needs to be specified.

### dplyr

in all commands always dataset - pipe - command

* %>%: command for Pipes. Useful for a series of filters or groups without making new variables
* summarize(): explore/ summarize the data
* group_by(): splits the dataset into hidden, internal groups based on one or more variables -> allows to perform group specific operations in the next step. often used in combination with summarize (group_by %>% summarize())
to ungroup, use ungroup() for following calaculations
* count(): see how many counts for each group exists (sort = TRUE arranges in descending order)
* filter(): extract rows from the data that meet a certain logical criteria
* distinct(): gives the distinct/ unique values for the selected variable (select multiple columns to get the combination for different variables)
* arrange(): sort and reorder rows in a dataset based on their values (sorts in an ascending order by default; for a descending order, add desc(value))
* select(): select the columns by name. The output is a tibble. the same function can be used for deselecting a column (just add an "!" in front of the columnName)
helper functions for select(): starts_with(""); ends_with(""); contains(""); num_range("")
* mutate(): create new columns based off other columns (e.g. calculating with two columns; with .before = columnName the new column will be added on the left side of the chosen column; the opposite with .after = columnName)

# Task Two
## Renewing Task 2 from last week
```{r}
library(tidyverse)

counts_df <- read_csv("basic_yeast_heatshock.csv")
gtf <- read_tsv("annotation.gtf", col_names = FALSE, comment = "#")

genes_filtered <- gtf %>% filter(X3 == "gene") %>% mutate(length = X5 - X4 + 1) # filter and calculate the gene length
lengths_table <- genes_filtered %>% mutate(gene_id = str_extract(X9, 'gene_id "[^"]+"') %>% str_remove_all('gene_id "|"')) %>% select(gene_id, length)
head(lengths_table)

counts_with_length <- counts_df %>% inner_join(lengths_table, by = "gene_id")

count_matrix <- counts_with_length %>% select(!length) %>% column_to_rownames("gene_id") %>% as.matrix()

genes_length <- counts_with_length$length
library_sizes <- colSums(count_matrix)
cpm_matrix <- t(t(count_matrix) / library_sizes) * 1000000
rpkm_matrix <- (count_matrix * 10^9) / (genes_length * rep(library_sizes, each = nrow(count_matrix)))

head(cpm_matrix)
head(rpkm_matrix)
                                          
```

## Renewing Task 3 from last week
```{r}


