---
title: "Assignment3"
output: html_document
---

### Basic Plotting in R

Task 1 a)
# One Dimensional Plotting
First we look at the possibilities to plot in R with one dimension:

Histogram
1D, many observations, ordered, distribution
```{r}
# here we plot the fuel efficiency in miles per gallon for each of the 32 cars in the dataset.

hist(mtcars$mpg)# mtcarsmpg is a build in dataset
```
Stem and leaf plot
Shows the mpg values of the 32 cars as a text based plot where each value is split into a stem (tens digit) and a leaf (ones digit).
1D, continous/discrete, few observations, orderd, distribution
```{r}
stem(mtcars$mpg)
```
Boxplot
1D, continuous, many observations, orderd by quantile
```{r}
boxplot(mpg ~ cyl, data = mtcars)#Plots one boxplot of mpg for each cylinder group (4, 6, 8), showing median, quartiles, and outliers of fuel efficiency per engine size.
```
Dot plot
1D, continuous, distribution
```{r}
stripchart(mtcars$mpg, method = "jitter")#Plots each car's mpg value as a single point along one axis, with jitter added so overlapping points are visible.
```
Barplot
1D, comparison/composition
```{r}
barplot(table(mtcars$cyl))#Plots the count of cars for each cylinder category (4, 6, 8) as bars.
```
Pie Chart
1D, categorical, few observations, unordered, compositions
```{r}
pie(table(mtcars$cyl))#Plots the share of cars in each cylinder category (4, 6, 8) as slices of a circle.
```

# Two Dimensional plotting

generic scatter plot
2D, continuous, few-many-observations
```{r}
plot(mtcars$wt, mtcars$mpg)#Plots each car as a point with weight on the x-axis and mpg on the y-axis to reveal how fuel efficiency relates to vehicle weight.
```

density shaded scatter
2D, continuous, many observations, unordered, distribution/relationship
```{r}
smoothScatter(rnorm(5000), rnorm(5000))#Plots 5000 randomly generated (x, y) points from a normal distribution as a smoothed color density, with darker areas indicating where many points overlap.
```

Scatter with overlap markers
```{r}
sunflowerplot(iris$Sepal.Length, iris$Sepal.Width)#Plots iris flowers as points with sepal length vs. sepal width, drawing "petals" on any point that represents multiple identical observations
```
quantile quantile plots
2D, continuous, many observations
```{r}
qqnorm(mtcars$mpg); qqline(mtcars$mpg)#Plots the quantiles of mpg against the theoretical quantiles of a normal distribution; the closer the points hug the reference line, the more normally distributed mpg is.
```

Multiple series on shared axes
2D, continuous, many observations, ordered 
```{r}
matplot(EuStockMarkets, type = "l")#Plots the daily closing prices of four European stock indices (DAX, SMI, CAC, FTSE) as four lines on the same chart over time.
```


spine plot
2D, categorical, ordered
```{r}
spineplot(Species ~ Sepal.Length, data = iris)#Plots how the proportions of the three iris species change across bins of sepal length, with bar widths scaled to bin frequency.
```
conditional density plot
```{r}
cdplot(Species ~ Sepal.Length, data = iris)#Plots a smoothed estimate of how the conditional probability of each iris species changes as sepal length increases
```

# Plotting in 3D

heatmap of a matrix
2D representation of 3D, continuous, many observations, ordered grid
```{r}
image(volcano)#Plots elevation values from the volcano dataset (Maunga Whau in Auckland) as a color-coded grid, where color indicates height.
```

contour lines
```{r}
contour(volcano)
```

perspective surface plot
3D, continuous, many observations, ordered grid, relationship

```{r}
persp(volcano, theta = 30, phi = 30)#Plots the volcano elevation data as a 3D wireframe surface viewed from a rotation angle of 30° and an elevation angle of 30°.

```
# Multidimensional Plotting

scatterplot matrix
multidimensional, unordered, continuous, many observations, relationship
```{r}
pairs(iris[, 1:4])#Plots a grid of scatter plots showing every pairwise combination of the four numeric iris measurements (sepal length, sepal width, petal length, petal width).
```


conditioning plot
multi dimensional, continuous, orderd conditioning
```{r}
coplot(mpg ~ wt | cyl, data = mtcars)#Plots mpg against weight as separate scatter panels, one for each cylinder group, to see if the weight-vs-mpg relationship differs by engine size.
```

mosaic plot for contingency tables
multi dimensional, categorical, many observations, unordered, composition/relationship
```{r}
mosaicplot(HairEyeColor)#Plots the joint distribution of hair color, eye color, and sex as nested rectangles whose areas are proportional to the cell frequencies in the contingency table.
```

# Task 1 b
We can write the labels for the x and y axis in the function.
The key words are xlab and ylab.
```{r}
plot(mtcars$wt, mtcars$mpg,
     xlab = "Weight (1000 lbs)",
     ylab = "Fuel efficiency (miles per gallon)")
```
We can use a log transformation by saying so in the function head:
```{r}
plot(mtcars$wt, log(mtcars$mpg),
     xlab = "Weight (1000 lbs)",
     ylab = "log(Fuel efficiency [mpg])")
```

Or we tell the function which axis should get log transformed:
```{r}
plot(mtcars$wt, mtcars$mpg, log = "y",
     xlab = "Weight (1000 lbs)",
     ylab = "Fuel efficiency (mpg, log scale)")
```
cex ("character expansion") scales text and points. 1 = default, 1.5 = 50 % larger, 0.8 = smaller.
```{r}
plot(mtcars$wt, mtcars$mpg,
     main = "Fuel efficiency vs. weight",
     xlab = "Weight (1000 lbs)", ylab = "MPG",
     cex      = 1.3,   # point size
     cex.main = 1.5,   # title size
     cex.lab  = 1.2,   # axis label size
     cex.axis = 0.9)   # tick number size
```


# Task 2

```{r}
raw_counts <- read.csv("basic_yeast_heatshock.csv", row.names = 1) # row.names = 1 moves 'gene_id' into the row labels position -> easy searching later on and no math Errors
log_counts <- log2(raw_counts + 1) # pseudocount for log-transformation

# Mean and standard deviation over all 9 replicates
genes_mean <- rowMeans(log_counts)
genes_sds <- apply(log_counts, 1, sd)

plot(genes_mean, genes_sds, main = "Gene Expression: Mean vs. Standar Deviation", xlab = "Mean Log2 Expression", ylab = "Standard Deviation", col = rgb(0,0,0,0.3), las = 1)
```

The scatter plot maps the global relationship between the average expression level of each gene and its standard deviation (variability) across all samples of the yeast heatshock experiment. The data reveals a high variability at an intermediate expression level, stabalizing at low and high Expression levels. the outlying points represent volatile genes with exceptionally high standard deviation, which undergo a condition specific regulatory shift. 
The graphic shows thas the log-transformation worked across the entire dataset.


```{r}
yfl_data <- as.numeric(log_counts["YFL014W", ])
x_positions <- c(1, 1, 1, 2, 2, 2, 3, 3, 3) # assign positions to the replicates
plot(x_positions, yfl_data, main = "Expression of YFL014W across all Replicates", xlab = "Condition", ylab = "Log2 Expression Count", col = "orchid", pch = 19, xaxt = "n", las = 1)
axis(1, at = c(1, 2, 3), labels = c("0 min", "10 min", "30 min")) # adding condition labels and positions
```

This scatter plot displays the raw expression values of the gene YFL014W across all nine replicates, grouped by their condition. Visualizing the individual data points confirms that the biological replicates within each condition cluster tightly together, indicating high reproducibility and low technical noise in the dataset. A upward shift in log2 expression counts is visible as the temperature rises and the time progresses, resulting in higher expression rate for this gene.


```{r}
yfl_mean_0min <- mean(yfl_data[1:3])
yfl_mean_10min <- mean(yfl_data[4:6])
yfl_mean_30min <- mean(yfl_data[7:9])
yfl_means <- c(yfl_mean_0min, yfl_mean_10min, yfl_mean_30min)

plot(c(1, 2, 3), yfl_means, type = "b", main = "Mean Expression of YFL014W per Condition", xlab = "Condition", ylab = "Mean Log2 Expression Count", col = "orchid", pch = 19, lwd = 2, xaxt = "n", las = 1)
axis(1, at = c(1, 2, 3), labels = c("0 min", "10 min", "30 min"))
```

By plotting the condition averages of this gene, this line plot puts the raw replicate data into a clean directional trajectory. The upward slope of the connecting line demonstrates a increase in mean log2 expression over the heat shock experiment. This visual representation focuses the attention on the overall trend rather than between sample variance, making it immediately that YFL014W ist strongly upregulated in response to thermal stress. 


```{r}
# Calculate the mean Expression for every gene
mean_0min <- rowMeans(log_counts[, 1:3])
mean_10min <- rowMeans(log_counts[, 4:6])
mean_30min <- rowMeans(log_counts[, 7:9])

# Identify the genes always increasing or decreasing
always_increasing <- (mean_0min < mean_10min) & (mean_10min < mean_30min)
always_decreasing <- (mean_0min > mean_10min) & (mean_10min > mean_30min)

# Create a list of the gene names
genes_always_increasing <- rownames(log_counts)[always_increasing]
genes_always_decreasing <- rownames(log_counts)[always_decreasing]

# Custom color vector for these genes
gene_colors <- rep(rgb(0.7, 0.7, 0.7, 0.2), nrow(log_counts))
gene_colors[always_increasing] <- rgb(0.1, 0.5, 0.8, 0.6)
gene_colors[always_decreasing] <- rgb(0.9, 0.2, 0.2, 0.6)


# Plot1: 0min vs. 10min
plot(mean_0min, mean_10min, main = "Gene Expression Mean: 0 min vs. 10 min", xlab = "Mean Log2 Expression (0 min)", ylab = "Mean Log2 Expression (10 min)", col = gene_colors, pch = 16, las = 1)
abline(a = 0, b = 1, col = "black", lwd = 1.5)
legend("topleft", legend = c("Always Increasing", "Always Decreasing", "Other"), col = c(rgb(0.1, 0.5, 0.5, 0.6), rgb(0.9, 0.2, 0.2, 0.6), rgb(0.7, 0.7, 0.7, 0.4)), pch = 16, bty = "n")


#Plot2: 10 min vs. 30min
plot(mean_10min, mean_30min, main = "Gene Expression Mean: 10 min vs. 30 min", xlab = "Mean Log2 Expression (10 min)", ylab = "Mean Log2 Expression (30 min)", col = gene_colors, pch = 16, las = 1)
abline(a = 0, b = 1, col = "black", lwd = 1.5)
legend("topleft", legend = c("Always Increasing", "Always Decreasing", "Other"), col = c(rgb(0.1, 0.5, 0.5, 0.6), rgb(0.9, 0.2, 0.2, 0.6), rgb(0.7, 0.7, 0.7, 0.4)), pch = 16, bty = "n")

```

The first plot compares the global mean expression of the genes between the control group (0 min) and the initial heat shock phase (10 min). The vast majority of genes cluster along the diagonal line, indicating that a significant portion of the transcriptome remains stable during the early stress response. The filtering successfully highlights a distinct subset of highly dynamic genes: the blue dots lying above the line represent transcripts that undergo rapid up-regulation, while the red dots below the line represent immediate down-regulation. The enhanced density of the grey background points ensures that the global distribution remains clearly visible while allowing these responding genes to stand out.

The second plot captures the subsequent transition of global gene expression between the 10-minute and 30-minute heat shock intervals. In contrast to the previous time step, the highlighted "Always Increasing" (blue) genes continue to maintain their position strictly above the diagonal line, proving that their expression continuously accumulates over the entire duration of the experiment. The "Always Decreasing" (red) genes remain below the diagonal, demonstrating an ongoing, progressive suppression. This visualization confirms that the identified gene subsets are not experiencing a temporary fluctuation, but are instead part of a sustained, coordinated transcriptional program to adapt to prolonged thermal stress.




