---
title: "Assignment3"
output: html_document
---

### Basic Plotting in R

Task 1 a)
# One Dimensional Plotting
First we look at the possibilities to plot in R with one dimension:

Histogram
1D, many observations, ordered, distribution
```{r}
# here we plot the fuel efficiency in miles per gallon for each of the 32 cars in the dataset.

hist(mtcars$mpg)# mtcarsmpg is a build in dataset
```
Stem and leaf plot
Shows the mpg values of the 32 cars as a text based plot where each value is split into a stem (tens digit) and a leaf (ones digit).
1D, continous/discrete, few observations, orderd, distribution
```{r}
stem(mtcars$mpg)
```
Boxplot
1D, continuous, many observations, orderd by quantile
```{r}
boxplot(mpg ~ cyl, data = mtcars)#Plots one boxplot of mpg for each cylinder group (4, 6, 8), showing median, quartiles, and outliers of fuel efficiency per engine size.
```
Dot plot
1D, continuous, distribution
```{r}
stripchart(mtcars$mpg, method = "jitter")#Plots each car's mpg value as a single point along one axis, with jitter added so overlapping points are visible.
```
Barplot
1D, comparison/composition
```{r}
barplot(table(mtcars$cyl))#Plots the count of cars for each cylinder category (4, 6, 8) as bars.
```
Pie Chart
1D, categorical, few observations, unordered, compositions
```{r}
pie(table(mtcars$cyl))#Plots the share of cars in each cylinder category (4, 6, 8) as slices of a circle.
```

# Two Dimensional plotting

generic scatter plot
2D, continuous, few-many-observations
```{r}
plot(mtcars$wt, mtcars$mpg)#Plots each car as a point with weight on the x-axis and mpg on the y-axis to reveal how fuel efficiency relates to vehicle weight.
```

density shaded scatter
2D, continuous, many observations, unordered, distribution/relationship
```{r}
smoothScatter(rnorm(5000), rnorm(5000))#Plots 5000 randomly generated (x, y) points from a normal distribution as a smoothed color density, with darker areas indicating where many points overlap.
```

Scatter with overlap markers
```{r}
sunflowerplot(iris$Sepal.Length, iris$Sepal.Width)#Plots iris flowers as points with sepal length vs. sepal width, drawing "petals" on any point that represents multiple identical observations
```
quantile quantile plots
2D, continuous, many observations
```{r}
qqnorm(mtcars$mpg); qqline(mtcars$mpg)#Plots the quantiles of mpg against the theoretical quantiles of a normal distribution; the closer the points hug the reference line, the more normally distributed mpg is.
```

Multiple series on shared axes
2D, continuous, many observations, ordered 
```{r}
matplot(EuStockMarkets, type = "l")#Plots the daily closing prices of four European stock indices (DAX, SMI, CAC, FTSE) as four lines on the same chart over time.
```


spine plot
2D, categorical, ordered
```{r}
spineplot(Species ~ Sepal.Length, data = iris)#Plots how the proportions of the three iris species change across bins of sepal length, with bar widths scaled to bin frequency.
```
conditional density plot
```{r}
cdplot(Species ~ Sepal.Length, data = iris)#Plots a smoothed estimate of how the conditional probability of each iris species changes as sepal length increases
```

# Plotting in 3D

heatmap of a matrix
2D representation of 3D, continuous, many observations, ordered grid
```{r}
image(volcano)#Plots elevation values from the volcano dataset (Maunga Whau in Auckland) as a color-coded grid, where color indicates height.
```

contour lines
```{r}
contour(volcano)
```

perspective surface plot
3D, continuous, many observations, ordered grid, relationship

```{r}
persp(volcano, theta = 30, phi = 30)#Plots the volcano elevation data as a 3D wireframe surface viewed from a rotation angle of 30° and an elevation angle of 30°.

```
# Multidimensional Plotting

scatterplot matrix
multidimensional, unordered, continuous, many observations, relationship
```{r}
pairs(iris[, 1:4])#Plots a grid of scatter plots showing every pairwise combination of the four numeric iris measurements (sepal length, sepal width, petal length, petal width).
```


conditioning plot
multi dimensional, continuous, orderd conditioning
```{r}
coplot(mpg ~ wt | cyl, data = mtcars)#Plots mpg against weight as separate scatter panels, one for each cylinder group, to see if the weight-vs-mpg relationship differs by engine size.
```

mosaic plot for contingency tables
multi dimensional, categorical, many observations, unordered, composition/relationship
```{r}
mosaicplot(HairEyeColor)#Plots the joint distribution of hair color, eye color, and sex as nested rectangles whose areas are proportional to the cell frequencies in the contingency table.
```

# Task 1 b
We can write the labels for the x and y axis in the function.
The key words are xlab and ylab.
```{r}
plot(mtcars$wt, mtcars$mpg,
     xlab = "Weight (1000 lbs)",
     ylab = "Fuel efficiency (miles per gallon)")
```
We can use a log transformation by saying so in the function head:
```{r}
plot(mtcars$wt, log(mtcars$mpg),
     xlab = "Weight (1000 lbs)",
     ylab = "log(Fuel efficiency [mpg])")
```

Or we tell the function which axis should get log transformed:
```{r}
plot(mtcars$wt, mtcars$mpg, log = "y",
     xlab = "Weight (1000 lbs)",
     ylab = "Fuel efficiency (mpg, log scale)")
```
cex ("character expansion") scales text and points. 1 = default, 1.5 = 50 % larger, 0.8 = smaller.
```{r}
plot(mtcars$wt, mtcars$mpg,
     main = "Fuel efficiency vs. weight",
     xlab = "Weight (1000 lbs)", ylab = "MPG",
     cex      = 1.3,   # point size
     cex.main = 1.5,   # title size
     cex.lab  = 1.2,   # axis label size
     cex.axis = 0.9)   # tick number size
```

