---
title: "MSMB Chapter 6 – Testing"
subtitle: "Key Concepts Explained"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

# Task 1

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 4, comment = "#>")
library(ggplot2)
library(dplyr)
library(tibble)
```

---

# 1 Null Hypothesis and Alternative Hypothesis

## The basic idea

When we analyse data, we usually want to know whether something interesting is
going on — for example, whether a gene is differentially expressed between two
conditions. But "interesting" needs a reference point. That reference is the
**null hypothesis**: it describes a boring, default world where nothing is
happening. In the gene expression example, the null hypothesis would be that the
gene's expression is the same in both conditions.

The **alternative hypothesis** is simply everything that is not the null — so any
situation where there actually *is* a difference. We never test the alternative
directly. Instead, we ask: how likely are our data if the null were true? If the
answer is "very unlikely", we reject the null and accept that the alternative must
hold.

## Why the alternative is true when the null fails

This works because together, null and alternative are exhaustive — they cover all
possibilities. The logic is: we set up a default assumption (null), and we only
abandon it when the data give us strong enough reason to. If the data are
incompatible with the null (low p-value), the null gets rejected, and by
exclusion, the alternative is what remains. This is not the same as proving the
alternative — it just means the null is no longer a reasonable description of
reality.

```{r null-viz, fig.height=3.5}
# Visualise the idea: null distribution and where our observation falls
x      <- seq(-4, 6, length.out = 500)
df_viz <- tibble(
  x    = x,
  null = dnorm(x, mean = 0, sd = 1)
)

obs <- 2.8   # our test statistic

ggplot(df_viz, aes(x, null)) +
  geom_area(data = filter(df_viz, x >= obs),
            fill = "#E41A1C", alpha = 0.4) +
  geom_line(linewidth = 1.1, colour = "steelblue") +
  geom_vline(xintercept = obs, linetype = "dashed") +
  annotate("text", x = obs + 0.15, y = 0.35,
           label = "observed\nstatistic", hjust = 0, size = 3.5) +
  annotate("text", x = 3.5, y = 0.05,
           label = "p-value\n(red area)", colour = "#E41A1C", size = 3.5) +
  labs(title = "Null distribution with p-value highlighted",
       x = "test statistic", y = "density") +
  theme_minimal()
```

---

# 2 The p-value

## What it is

The p-value is the probability of observing a test statistic at least as extreme
as the one we actually got — *assuming the null hypothesis is true*. Visually, it
is the red area in the plot above: the tail of the null distribution beyond our
observation.

A small p-value means our data would be very unusual in a world where the null
is true. We use this as evidence against the null.

## What it is NOT

The p-value is frequently misunderstood. It is **not**:

- the probability that the null hypothesis is true
- the probability that our result occurred by chance
- a measure of effect size or biological relevance

A very large dataset can produce a tiny p-value for a biologically meaningless
difference. And a large p-value does not prove the null is true — it just means
we don't have enough evidence to reject it.

```{r pvalue-demo}
# Coin tossing example from the book
set.seed(42)
numFlips <- 100
numHeads <- 59

# Null distribution: binomial with p=0.5
k             <- 0:numFlips
binomDensity  <- tibble(k = k, p = dbinom(k, size = numFlips, prob = 0.5))
alpha         <- 0.05
binomDensity  <- arrange(binomDensity, p) |>
                   mutate(reject = cumsum(p) <= alpha)

ggplot(binomDensity) +
  geom_bar(aes(x = k, y = p, fill = reject), stat = "identity") +
  scale_fill_manual(values = c("TRUE" = "#E41A1C", "FALSE" = "grey70"),
                    labels = c("TRUE" = "rejection region", "FALSE" = "no rejection")) +
  geom_vline(xintercept = numHeads, colour = "steelblue", linewidth = 1) +
  annotate("text", x = numHeads + 1, y = 0.07,
           label = paste("observed:", numHeads), hjust = 0, colour = "steelblue") +
  labs(title = "Binomial null distribution for 100 coin flips (p = 0.5)",
       subtitle = "Red = rejection region at α = 0.05 | Blue line = observed value",
       x = "number of heads", y = "probability", fill = "") +
  theme_minimal()

# Exact p-value
pval <- binom.test(numHeads, numFlips, p = 0.5)$p.value
cat(sprintf("p-value: %.4f  →  %s reject at α = 0.05\n",
            pval, ifelse(pval < 0.05, "DO", "do NOT")))
```

> The observed value of 59 heads sits in the grey region — we do not reject the
> null. Despite knowing (from the simulation) that the coin is biased (p=0.6),
> 100 flips are not enough to prove it with 95% confidence.

---

# 3 Types of Error

When we make a decision based on a test, we can be right or wrong in two
different ways:

| | **Null is actually true** | **Null is actually false** |
|---|---|---|
| **We reject null** | Type I error (false positive) | True positive ✓ |
| **We don't reject** | True negative ✓ | Type II error (false negative) |

## Type I error (false positive)

We conclude there is an effect when there isn't one. The rate at which this
happens is $\alpha$, which we set ourselves (usually 0.05). This is also called
the **significance level**.

## Type II error (false negative)

We miss a real effect. The rate is $\beta$. The **power** of a test is
$1 - \beta$ — how often it correctly detects a true effect.

## The trade-off

You can never eliminate both errors at once. Making the test stricter (lower
$\alpha$, shifting the threshold) reduces false positives but inevitably
increases false negatives — and vice versa. The right balance depends on the
context: missing a real drug effect (false negative) and falsely approving an
ineffective drug (false positive) have very different consequences.

```{r error-types, fig.height=4}
x    <- seq(-4, 8, length.out = 1000)
thresh <- 1.96

df_errors <- tibble(
  x   = x,
  H0  = dnorm(x, mean = 0, sd = 1),
  H1  = dnorm(x, mean = 3, sd = 1)
)

ggplot(df_errors, aes(x)) +
  geom_area(data = filter(df_errors, x >= thresh),
            aes(y = H0), fill = "#E41A1C", alpha = 0.5) +
  geom_area(data = filter(df_errors, x <= thresh),
            aes(y = H1), fill = "#377EB8", alpha = 0.5) +
  geom_line(aes(y = H0), colour = "#E41A1C", linewidth = 1) +
  geom_line(aes(y = H1), colour = "#377EB8", linewidth = 1) +
  geom_vline(xintercept = thresh, linetype = "dashed") +
  annotate("text", x = 3.2,  y = 0.32, label = "H₁ (alternative)",
           colour = "#377EB8", size = 3.5) +
  annotate("text", x = -1.5, y = 0.32, label = "H₀ (null)",
           colour = "#E41A1C", size = 3.5) +
  annotate("text", x = 2.7,  y = 0.05, label = "Type I\n(α)",
           colour = "#E41A1C", size = 3) +
  annotate("text", x = 0.8,  y = 0.05, label = "Type II\n(β)",
           colour = "#377EB8", size = 3) +
  labs(title = "Type I and Type II errors",
       subtitle = "Dashed = decision threshold",
       x = "test statistic", y = "density") +
  theme_minimal()
```

---

# 4 Multiple Testing

## The problem

In biology we rarely test just one hypothesis. A typical RNA-Seq experiment tests
thousands of genes simultaneously. If we use $\alpha = 0.05$ for each test, we
expect $5\%$ of all tests to produce a false positive — even when nothing is
truly differentially expressed. For 20,000 genes, that's 1,000 false positives
by chance alone.

The xkcd comic the book references illustrates this perfectly: if you test 20
jelly bean colours for causing acne, one of them will likely come up "significant"
purely by chance — and that's the one that gets reported in the newspaper.

## Family-wise error rate (FWER) and Bonferroni

One approach is to control the **family-wise error rate**: the probability of
making *at least one* false positive across all tests. The Bonferroni correction
divides $\alpha$ by the number of tests $m$:

$$\alpha_{\text{per test}} = \frac{\alpha_{\text{FWER}}}{m}$$

This is very conservative — with 20,000 genes, each individual test needs to
reach $p < 0.0000025$ to be called significant. Real effects can easily be
missed.

```{r bonferroni-demo}
m     <- 20000
alpha <- 0.05
cat("Bonferroni threshold per test:", alpha / m, "\n")

# How FWER blows up without correction
tibble(m_tests = c(1, 10, 100, 1000, 10000),
       FWER    = 1 - (1 - alpha)^m_tests) |>
  knitr::kable(digits = 4,
               col.names = c("Number of tests", "FWER (no correction)"))
```

## Multiple testing as an opportunity

The book makes an important point: multiple testing is not just a burden. When
you have thousands of p-values, you can *learn from their distribution*. The
histogram of p-values is itself informative — it tells you how many true effects
there are and how strong they are. This is something you can never get from a
single test.

---

# 5 False Discovery Rate

## A better question

Instead of asking "will I make any false positive at all?" (FWER), the FDR asks
a more practical question: **among all the things I call significant, what
fraction are actually false positives?**

$$\text{FDR} = \mathbb{E}\!\left[\frac{\text{false positives}}{\text{all rejections}}\right]$$

If I reject 100 hypotheses and control FDR at 10%, I accept that about 10 of
those 100 are probably wrong. For a genome-wide screen where we're going to
follow up anyway, this is a completely reasonable trade-off.

## The p-value histogram

Before applying any FDR method, always plot the histogram of all p-values. It
should show:

- a **peak near 0** — these are the true positives (genes with real effects)
- a **flat uniform part** — these are the true nulls (p-values from null hypotheses
  are uniformly distributed between 0 and 1 by definition)

```{r pvalue-histogram}
set.seed(99)
m        <- 5000
m0       <- 4000   # true nulls
m1       <- m - m0 # true alternatives

p_null <- runif(m0)                        # nulls: uniform
p_alt  <- rbeta(m1, shape1 = 0.3, shape2 = 1)  # alternatives: enriched near 0

p_all  <- c(p_null, p_alt)

ggplot(tibble(p = p_all), aes(x = p)) +
  geom_histogram(binwidth = 0.05, boundary = 0,
                 fill = "steelblue", colour = "white") +
  geom_hline(yintercept = m0 * 0.05,   # expected null count per bin
             colour = "#E41A1C", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 0.6, y = m0 * 0.05 + 15,
           label = "expected null level", colour = "#E41A1C", size = 3.5) +
  labs(title = "Simulated p-value histogram",
       subtitle = paste0(m1, " true alternatives + ", m0, " true nulls"),
       x = "p-value", y = "count") +
  theme_minimal()
```

## Benjamini-Hochberg procedure

The most widely used method for FDR control. The steps:

1. Sort all $m$ p-values: $p_{(1)} \leq p_{(2)} \leq \cdots \leq p_{(m)}$
2. Choose target FDR level $\varphi$ (e.g. 0.05 or 0.10)
3. Find the largest $k$ such that $p_{(k)} \leq \varphi \cdot k / m$
4. Reject all hypotheses $1, \ldots, k$

```{r bh-procedure, fig.height=4}
phi   <- 0.10
p_sorted <- sort(p_all)
m_tot    <- length(p_sorted)
rank_vec <- seq_along(p_sorted)

# Find BH cutoff
bh_line  <- phi * rank_vec / m_tot
k_max    <- max(which(p_sorted <= bh_line))
p_cutoff <- p_sorted[k_max]

cat(sprintf("BH threshold at FDR = %.2f: p ≤ %.4f → %d rejections\n",
            phi, p_cutoff, k_max))

# Visualise only the smallest 1000 p-values for clarity
n_show <- 1000
ggplot(tibble(rank = rank_vec[1:n_show], p = p_sorted[1:n_show]),
       aes(rank, p)) +
  geom_line(linewidth = 0.8) +
  geom_abline(slope = phi / m_tot, intercept = 0,
              colour = "#E41A1C", linewidth = 1) +
  geom_vline(xintercept = k_max, linetype = "dashed", colour = "#4DAF4A") +
  annotate("text", x = k_max + 30, y = 0.006,
           label = paste("reject\nup to here\n(k =", k_max, ")"),
           hjust = 0, colour = "#4DAF4A", size = 3) +
  labs(title = "Benjamini-Hochberg procedure",
       subtitle = paste0("Red line = slope φ/m = ", round(phi/m_tot, 6),
                         "  |  Reject where p-value line is BELOW red line"),
       x = "rank", y = "p-value") +
  theme_minimal()
```

## Correction method choice: Bonferroni vs. BH

| | **Bonferroni** | **Benjamini-Hochberg** |
|---|---|---|
| Controls | FWER (≥1 false positive) | FDR (fraction false positives) |
| How strict | Very conservative | More lenient |
| Best for | Confirmatory, few tests, high stakes | Exploratory, many tests |
| Risk | Many missed real effects | Some false positives in the hit list |

**My choice:** For a genome-wide screen (like the yeast heatshock data), BH is
the right approach. We're looking for candidates to follow up, not making final
claims — a small fraction of false positives in the hit list is acceptable, and
Bonferroni would throw away too many real hits.

```{r correction-comparison}
# Direct comparison on the same data
p_bonferroni <- p.adjust(p_all, method = "bonferroni")
p_bh         <- p.adjust(p_all, method = "BH")

cat("Rejections at α = 0.05:\n")
cat("  Bonferroni:", sum(p_bonferroni < 0.05), "\n")
cat("  BH        :", sum(p_bh         < 0.05), "\n")
cat("  Truth     :", m1, "true alternatives\n")
```

---

# 6 Summary

| Concept | In one sentence |
|---|---|
| **Null hypothesis** | The default assumption of "nothing happening"; we test against it |
| **Alternative hypothesis** | Everything that is not the null; accepted when null is rejected |
| **p-value** | Probability of data at least this extreme *if* the null were true |
| **Type I error** | False positive — rejecting a true null (rate = α) |
| **Type II error** | False negative — missing a real effect (rate = β) |
| **Multiple testing** | Running many tests inflates false positives; needs correction |
| **Bonferroni** | Divide α by number of tests; very strict FWER control |
| **FDR** | Expected fraction of false positives among all rejections |
| **BH procedure** | Practical FDR control; sorts p-values and finds a cutoff |
| **p-value histogram** | Essential sanity check; flat = mostly nulls, peak at 0 = real signal |


# Task 2

```{r}
library(tidyverse)
library(DESeq2)

raw_data <- read_csv("basic_yeast_heatshock.csv")

counts_10min <- raw_data %>% select(gene_id, `25_0_r1`, `25_0_r2`, `25_0_r3`, `42_10_r1`, `42_10_r2`, `42_10_r3`)  %>% column_to_rownames(var = "gene_id") %>% as.matrix()

col_data_10min <- data.frame(condition = factor(c("control_0min", "control_0min", "control_0min", "heat_10min",   "heat_10min",   "heat_10min")))
rownames(col_data_10min) <- colnames(counts_10min)

col_data_10min$condition <- relevel(col_data_10min$condition, ref = "control_0min") # control group as baseline

all(colnames(counts_10min) == rownames(col_data_10min))

dds_10min <- DESeqDataSetFromMatrix(countData = counts_10min,
                              colData = col_data_10min,
                              design = ~ condition)

dds_10min <- DESeq(dds_10min)

result_10min <- results(dds_10min, contrast = c("condition", "heat_10min", "control_0min"))

res_10min_df <- result_10min %>% as.data.frame() %>% rownames_to_column(var = "gene_id")

head(res_10min_df)

counts_30min <- raw_data %>% select(gene_id, `25_0_r1`, `25_0_r2`, `25_0_r3`, `42_30_r1`, `42_30_r2`, `42_30_r3`) %>% column_to_rownames(var = "gene_id") %>% as.matrix()

col_data_30min <- data.frame(condition = factor(c("control_0min", "control_0min", "control_0min",
    "heat_30min",   "heat_30min",   "heat_30min")))
rownames(col_data_30min) <- colnames(counts_30min)
col_data_30min$condition <- relevel(col_data_30min$condition, ref = "control_0min")
all(colnames(counts_30min) == rownames(col_data_30min))

dds_30min <- DESeqDataSetFromMatrix(countData = counts_30min,
                                    colData = col_data_30min,
                                    design = ~ condition)

dds_30min <- DESeq(dds_30min)

result_30min <- results(dds_30min, contrast = c("condition", "heat_30min", "control_0min"))
res_30min_df <- result_30min %>% as.data.frame() %>% rownames_to_column(var = "gene_id")

head(res_30min_df)

up_10min <- res_10min_df %>% filter(padj < 0.05 & log2FoldChange > 1)
down_10min <- res_10min_df %>% filter(padj < 0.05 & log2FoldChange < -1)

up_30min <- res_30min_df %>% filter(padj < 0.05 & log2FoldChange > 1)
down_30min <- res_30min_df %>% filter(padj < 0.05 & log2FoldChange < -1)

```

```{r - volcano plot}
plot_data_10min <- res_10min_df %>% 
  mutate(Category = case_when(padj < 0.05 & log2FoldChange > 1 ~ "Upregulated", 
                               padj < 0.05 & log2FoldChange < -1 ~ "Downregulated", 
                               TRUE ~ "Not significant"))

ggplot(plot_data_10min, aes(x = log2FoldChange, y = -log10(padj), color = Category)) + geom_point(alpha = 0.5, size = 1.5) + scale_color_manual(values = c("Upregulated" = "red",
                                                    "Downregulated" = "blue",
                                                    "Not significant" = "grey")) + 
  theme_minimal() + 
  labs(title = "Volcano Plot: Heatshock 10 min vs. control", x = "Log2 Fold Change", y = "-Log10 adjusted p-value")
```

```{r - bar plot}
data_barplot <- data.frame(Time = c("10 min", "10 min", "30 min", "30 min"),
                           Direction = c("Upregulated", "Downregulated", "Upregulated", "Downregulated"),
                           Number = c(nrow(up_10min), nrow(down_10min), nrow(up_30min), nrow(down_30min)))
                           
ggplot(data_barplot, aes( x = Time, y = Number, fill = Direction)) + geom_bar(stat = "identity", position = "dodge") + scale_fill_manual(values = c("Upregulated" = "red", "Downregulated" = "blue")) + theme_bw() + 
  labs(title = "Comparison of the Number expressed genes",
       x = "Time of Heatshock",
       y = "Amount of genes (p_adj < 0.05, |log2FoldChange| > 1)",
       fill = "Regulation") + 
  geom_text(aes(label = Number), position = position_dodge(width = 0.9), vjust = -0.25)
```


```{r - intercept}
u10 <- up_10min %>% select(gene_id) %>% mutate(up_10min = TRUE)
d10 <- down_10min %>% select(gene_id) %>% mutate(down_10min = TRUE)
u30 <- up_30min %>% select(gene_id) %>% mutate(up_30min = TRUE)
d30 <- down_30min %>% select(gene_id) %>% mutate(down_30min = TRUE)

overlap_matrix <- u10 %>% 
  full_join(d10, by = "gene_id") %>%
  full_join(u30, by = "gene_id") %>%
  full_join(d30, by = "gene_id")

overlap_clean <- overlap_matrix %>%
  replace_na(list(up_10min = FALSE, down_10min = FALSE, up_30min = FALSE, down_30min = FALSE)) # NA -> FALSE

overlap_counts <- overlap_clean %>% dplyr::count(up_10min, down_10min, up_30min, down_30min, name = "Number_Genes") 

overlap_long <- overlap_counts %>% mutate(Combination_ID = paste0("Intercept_", row_number())) %>% pivot_longer(cols = c(up_10min, down_10min, up_30min, down_30min), names_to = "Gen_Set", values_to = "is_valid")

ggplot(overlap_long, aes(x = reorder(Combination_ID, -Number_Genes), y = Gen_Set)) + 
  geom_tile(aes(fill = is_valid), color = "white", size = 1) + 
  scale_fill_manual(values = c("TRUE" = "orchid", "FALSE" = "skyblue")) + geom_text(aes(label = ifelse(is_valid, as.character(Number_Genes), "")), color = "white", fontface = "bold") + 
  theme_minimal() + 
  labs(title = "Intercept of the 4 Gene Sets",
       x = "Combination of Intercept (sorted by number)",
       y = "Gene Set",
       fill = "Contained in intercept") + 
  theme(axis.text.x = element_blank(),
        panel.grid = element_blank())
  
