---
title: "GeneAnnotation"
output: html_document
---

# Chapter One: Generative Models for Discrete Data

```{r - Poisson Distribution}
#for one value
dpois(x = 3, lambda = 5)

#for several values instead of a loop
dpois(x = 0:12, lambda = 5)
barplot(dpois(0:12, 5), names.arg = 0:12, col = "red")
```
The first command means that the probability for 3 mutations in the genome after one replication cycle is 0.14 percent

The Second code shows the probability for 0-12 mutations after one replication cycle in a barplot.

## Using discrete probability models

```{r - factors}

genotype = c("AA","AO","BB","AO","OO","AO","AA","BO","BO","AO","BB","AO","BO","AB","OO","AB","BB","AO","AO")
table(genotype)

#with factor:
genotypeF =  factor(genotype)
levels(genotypeF)
table(genotypeF)
```
with factors, R automatically detects all the layers of a function and encodes the categorial variables. Levels define which values are allowed in a set of data.

### Bernoulli trials

```{r - binomial r-funktion}
# r-functions generate tailored outcomes for each type of distribution
rbinom(15, prob = 0.5, size = 1)

# withs unequal probabilities
rbinom(12, prob = 2/3, size = 1)
```
**Parameters:**
Number of Trials, probability of success, repetitions per trial

This code chunk represents binomial experiment. The first a coin toss with equal probability of head and tail, the second one a ball drop into boxes with different sizes and therefore unequal probabilities. 

```{r}
#with changed parameters
rbinom(1, prob = 2/3, size = 12)
```
**Parameters**:
Number of successes, probability of success, Number of Trials

This command counts the number of successful events in the same experiment as the second above (with the unequal probabilities). The number of unsuccessful events can be calculated as Number of Trials - Number of successes.

### Binomial success counts

```{r}
set.seed(235569515)
rbinom(1, prob = 0.3, size = 15)
```
`set.seed` determines a starting point for the trials, so that the outcome always stays the same. This is important to reproduce experiments (reliability).

```{r}
probabilities = dbinom(0:15, prob = 0.3, size = 15)
round(probabilities, 2)
barplot(probabilities, names.arg = 0:15, col = "red")
```
`round()`controls the number of printed decimal digits
`names.arg`Labels the x-axis bars with the possible event outcomes

### Poisson distributions
```{r}
rbinom(1, prob = 5e-4, size = 10000)
simulations = rbinom(n = 30000, prob = 5e-4, size = 10000)
barplot(table(simulations), sol = "lavender")
```
using the poisson distribution, when Number of trials is large and the probability of success is small. 
The Parameter is Lambda

### A generative model for epitope detection
```{r, eval=FALSE}
load("../data/e100.RData")
barplot(e100, ylim = c(0, 7), width = 0.7, xlim = c(-0.5, 100.5),
  names.arg = seq_along(e100), col = "darkolivegreen")
```
**Parameter:**
e100: data which is plotted
ylim: sets y-axis
width: width of the bars
xlim: setzs x-axis
names.arg = seq_along(e100): generates Numbers from 1 to 100 and labels the x-axis

```{r}
1 - ppois(6, 0.5)
ppois(6, 0.5, lower.tail = FALSE)
```
Both commands yield the same result. ppois(x, lambda) calculates the probability of x or fewer results. By setting lower.tail = FALSE, the command calculates the probability for strictly more than x results (P(X > x)).

```{r - monte-carlo-simulation}
maxes = replicate(100000, {
  max(rpois(100, 0.5))
})
table(maxes)
```
**Parameter:**
replicate(x, {}): replicates the code inside x-times
max(rpois(100, 0.5)): takes the maximal value of 100 Poisson distributed numbers
This simulation shows the distribution of the maximum number of mutations found in groups of 100 viruses. While the average is low (0.5), observing 3 or 4 mutations as a group maximum is very common.

## Multinomial distribution: the case of DNA
```{r}
pvec = rep(1/4, 4)
t(rmultinom(1, prob = pvec, size = 8)) #rmultinom simulates categorical data
```
**Parameters:**
1/4: probability for success for each outcome
4: number of replications
rmultinom(n,  prob, size): n = number of independent simulations, prob = vector of probabilities (chance of each category), size = number of observations in each simulation
1: number of vectors to generate 
size: number if events
pvec creates a vector with the results of the multinomial distribution
t() transponses (flips) a matrix or dataframe

### Simulating for power
```{r}
obsunder0 = rmultinom(1000, prob = pvec, size = 20)
dim(obsunder0)

expected0 = pvec * 20
sum((obsunder0[, 1] - expected0)^2 / expected0)
```
dim() returns the rows and columns of a matrix
expected0: calculates the expected value
obsunder0[, n] - expected0: calculates the difference between the observed and expected value of all rows in the column n
/ expected0: standardization

```{r - Chi^2}
stat = function(obsvd, exptd = 20 * pvec) {
  sum((obsvd - exptd)^2 / exptd)
}
stat(obsunder0[, 1])

S0 = apply(obsunder0, 2, stat)
summary(S0)

hist(S0, breaks = 25, col = "lavender", main = "")

q95 = quantile(S0, probs = 0.95)
q95
```
apply(x, MARGIN, FUN): x = data source; MARGIN = direction (1 = rows, 2 = columns); FUN = applied function
summary(): shows statistical indicators of the function
hist(x, breaks, col, main): x = dataset, breaks = number of columns, col = folour, main = title
quantile(x, probs): x = numeric vector whose quantiles are wanted (dataset); probs = numeric vector of probabilities [0, 1]

```{r alternative Hypothesis}
pvecA = c(3/8, 1/4, 1/4, 1/8)
observed = rmultinom(1000, prob = pvecA, size = 20)
dim(observed)
apply(observed, 1, mean)
expectedA = pvecA * 20
expectedA
stat(observed[, 1])
S1 = apply(observed, 2, stat)
q95
sum(S1 > q95) #counts the successes of the simulation
power = mean(S1 > q95) #the proportion of simulations where the test statistic exceeded the critical value
power
```
An alternative Hypothesis for pvec, so calculation of power can be made
The same structur than before, just with power analysis at the end
sum(..., na.rm): ... = numerical, complex or logical vector, na.rm = should missing values be removed? (default = FALSE)
mean(x, na.rm, trim): x = the object (numerical or logical vector), na.rm = should missing values be removed? (default = FALSE), trim = cuts off data at the extreme of x (remove outliers)

# Chapter two: Statistical Modeling

## A simple example of statistial modeling
```{r eval=FALSE}
rootogram() # diagnostic plot to evaluate the goodness of fit for probability models with lots of different parameters
prod() # returns the product of all the values present in its arguments

```





