---
title: "Networks"
output: html_document
date: "2026-06-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task One
```{r - transcription factor and target genes}
library(tidyverse)
library(ggplot2)

network_colnames = c('Regulator_systematic', 'Regulator_standard', 'Target_systematic', 'Target_standard', 'Id', 'Date', 'Environment', 'Regulation', 'Type', 'Method')

yeast_network <- read_tsv("yeastract2019-flat-file.tsv", col_names = network_colnames)

tf_count <- yeast_network %>% distinct(Regulator_systematic) %>% nrow()

target_count <- yeast_network %>% distinct(Target_systematic) %>% nrow()

cat("Number of TF:", tf_count)
cat("number of target genes:", target_count)

nodes_data <- data.frame(
  Category = c("Transcription Factors", "Target Genes"),
  Count = c(tf_count, target_count))
  
  ggplot(nodes_data, aes(x = Category, y = Count, fill = Category)) +
    geom_bar(stat = "identity", color = "black", width = 0.5) +
    geom_text(aes(label = scales::comma(Count)), vjust = -0.5, color = "black", fontface = "bold", size = 5) + scale_fill_manual(values = c("Transcription Factors" = "purple2", "Target Genes" = "darkseagreen")) + 
    scale_y_continuous(labels = scales::comma) + 
    expand_limits(y = max(tf_count, target_count) * 1.15) +
    labs(
    title = "Total Number of Unique Network Entities",
    subtitle = "Comparison of Unique Regulatory Factors vs. Target Genes",
    x = "Network Component",
    y = "Count") + 
    theme_minimal() + 
    theme(legend.position = "none")
  


```


There are fare more Target genes compared to Transcription factors, which means that the Transcription Factors regulate lots of different genes.This can be seen later on in this analysis

```{r - edges}
total_edges <- nrow(yeast_network)

unique_pairs <- yeast_network %>% distinct(Regulator_systematic, Target_systematic) %>% nrow()

redundant_edges <- total_edges - unique_pairs

cat("Total numbers of edges:", total_edges)
cat("Number of unique regulator-target-pairs:", unique_pairs)
cat("Number of redundant regulator-target pairs:", redundant_edges)

plot_data <- data.frame(
  Type = c("Unique Pairs", "Redundant Edges"), Count = c(unique_pairs, redundant_edges))

ggplot(plot_data, aes(x = Type, y = Count, fill = Type)) + geom_bar(stat = "identity", color = "black", width = 0.6) + 
  geom_text(aes(labels = scales::comma(Count)), vjust = - 0.7, color = "black", fontface = "bold", size =  5) + 
  scale_fill_manual(values = c("Unique Pairs" = "skyblue", "Redundant Edges" = "salmon")) + 
  scale_y_continuous(labels = scales::comma) + 
  ylim(0, 220000) +
  labs(title = "Comparison of Uniqe ve. Reduntand Netword Edges", subtitle = paste("Total observed rows in File:", scales::comma(total_edges)), 
       x = "Edge Category",
       y = "Number of Edges") + 
  theme_minimal() + 
  theme(legend.position = "none")
```


```{r - TF and targets}
targets_per_tf <- yeast_network %>% group_by(Regulator_systematic) %>% summarize(number_targets = n_distinct(Target_systematic))

ggplot(targets_per_tf, aes(x = number_targets)) + geom_histogram(bins = 12, fill = "skyblue", color = "black") + labs(title = "Number of target Genes per Transkription Factor", x = "Number of regulated Genes", y = "number of Transcription Factor") + theme_minimal() + scale_y_continuous(breaks = seq(0, 100, by = 20))


tf_per_target <- yeast_network %>% group_by(Target_systematic) %>% summarize(number_tf = n_distinct(Regulator_systematic))

ggplot(tf_per_target, aes(x = number_tf)) + geom_histogram(bins = 12, fill = "salmon", color = "black") + labs(title = "Number of Transcription factors per Target Gene", x = "Number of Transxription Factors", y = "Number of target Genes") + theme_minimal()

```

The first histogram displays the number of unique target genes regulated by individual transcription factors (TFs), representing the network's out-degree distribution.

Scale-Free Property: The distribution exhibits a heavy-tailed, power-law-like shape. The vast majority of transcription factors regulate only a very small number of target genes (fewer than 500).

Network Hubs: Conversely, a selective handful of highly connected TFs act as global network "hubs," regulating an exceptionally high number of genes (some controlling over 6,000 targets). Biologically, these hubs represent master regulators responsible for fundamental cellular processes, global stress responses, or large-scale metabolic shifts.

The second histogram displays the number of distinct transcription factors regulating a single target gene, representing the network's in-degree distribution.

Combinatorial Regulation: Unlike the out-degree distribution, this profile is shifted to the right, showing a clear peak where the majority of target genes are regulated by approximately 20 to 40 different transcription factors simultaneously.

Biological Significance: This pattern is indicative of complex combinatorial regulation in eukaryotic systems. Most yeast genes do not rely on a simple on/off switch controlled by a single factor. Instead, their expression is fine-tuned via integrating inputs from multiple upstream regulators, allowing the cell to precisely coordinate gene expression based on complex environmental and physiological cues.

```{r - timeframe histogram}
timeframe <- yeast_network %>% mutate(Year = as.numeric(str_sub(Date, start = 1, end = 4)))

ggplot(timeframe, aes(x = Year)) + geom_histogram(binwidth = 1, fill = "lightgreen", color = "black") + labs(title = "Published Expression Data over Time", x = "Year", y = "Number of published data") + theme_minimal()
```

The timeframe of the histogram is between 1976 and 2019, with a peak in published data in 2008

```{r - distributions}
ggplot(yeast_network, aes(x = Type)) + geom_bar(fill = "plum", color = "black") + labs(title = "Distribution of Interaction type", x = "Type", y = "Number") + theme_minimal()

yeast_network %>% filter(!is.na(Method)) %>% mutate(Method_Clean = fct_lump_n(Method, n = 15)) %>% ggplot(aes(y = fct_infreq(Method_Clean))) + geom_bar(fill = "plum", color = "black") + labs(title = "Distribution of Top 15 Experimental Methods", x = "Number", y = "Experimental Method") + theme_minimal() + facet_wrap(~ Type)

ggplot(yeast_network, aes(x = Environment)) + geom_bar(fill = "plum", color = "black") + labs(title = "Distribution of environmental conditions", x = "Environmental Condition", y = "Number") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + facet_wrap(~ Regulation)

ggplot(yeast_network, aes(x = Regulation)) + geom_bar(fill = "plum", color = "black") + labs(title = "Distribution of Regulation Type", x = "Regulation Type", y = "Number") + theme_minimal() + facet_wrap(~ Type)


```


As shown in the plots above, there are some imbalances

**Experimental Methods faceted by Interaction Type:**
Chip-on-Chip and Chip-seq appear almost exclusively under the direct interaction facet. This is reasonable and backed by the biological background of the methods, using transcription factors as direct link for the target genes to identify the DNA binding sites
On the other side, Microarrays appear exclusively within the indirect interaction facet. Expression profiling via Microarrays captures downstream transcriptional changes (indirect regulatory cascade)

**Regulation Type split by Interaction Type:**
Direct Interactions are mostly N/A: For direct interactions (often derived from high-throughput ChIP-seq or ChIP-on-chip data), the mode of regulation is predominantly unknown (N/A). Binding assays prove that a protein sits on a piece of DNA, but they cannot inherently deduce whether that binding event activates or represses transcription.

Indirect Interactions specify Direction: For indirect interactions, the direction (Positive or Negative) is almost always known. This is because these edges are derived from expression-based assays (RNA-Seq, Microarrays, Northern Blots) where an increase or decrease in target mRNA levels is directly measurable.


