---
title: "task3"
output: html_document
---

```{r}
library(here)
sample <- read.csv2(here("data", "sample.tsv"), sep="\t", header=TRUE)

```
## Data Overview 1
```{r}
head(sample)
dim(sample)
sum(sample$count==0)
```

## Data Overview 2
```{r}
min(sample$count)
max(sample$count)
mean(sample$count)
sd(sample$count)
```
# Plotting
```{r}
plot(sample$count)
barplot(sample$count)
stripchart(sample$count)
boxplot(sample$count)
hist(sample$count)
plot(density(sample$count))
```
row 12: read.csv2 imports the data of the chosen file (sample.tsv); the parameters specify that the file uses tabs (sep="\t") as separators and contains column names in the first row (header = TRUE)
row 16: head() shows the first six rows of the dataset to provide a quick look into the data structure
row 17: dim() shows the dimensions of the daraset (number of rows and columns)
row 18: sum(sample$count==0) calculates the total number of genes without expression 
row 21 & 22: min()/ max() identify the lowest/ highest expression values int the column "count"
row 23: mean() calculates the average expression level
row 24: sd() calculates the standard deviation (shows dispersion around the average)
row 27-32: create 6 different plot as visual representation of the values

sneak peak: shows the first 6 genes and its amount of expression. For example, in these 6 genes, the highest expression is in gene ICR1 with 194 counts.
general statistics: Result of dim() with 7,126 rows (= genes) und 2 columns (gene_id and counts)
          1,090 genes have an expression value of zero 
          the counts range from a min of 0 to a max. 194,695 (no transciption under these circumstances)
          the mean is approximately 749, the sd 3837 (high sd indicates a high level of dispersion)
Indexplot: each point represents 1 gene. The x-axis could be the row number in the table, the y-axis the expression count
Most genes have low expression (under 30,000 counts) while few outliers are expressed highly (>100,000 counts)
Barplot: shows a expression as a bar for each gene in the dataset. its shows the same like the Indexplot, just a different type of graphics/ vizualisation.
Stripchart: is a one-dimensional scatter plot, that stacks the expression values on an horizontal axis. shows high density and gaps in the expression of genes of this dataset.
Boxplot: the box (not visible) represents the middle 50% of the data. looks like a line in the bottom part, because the mean and most values have similar expression rates
Histogram: gathers frequencies of different expression rates in columns. Each column represents 10,000 counts in expression rate. the first column contains almost all values
Density plot: is a smoothed version of the Histogram showing the probalility distribution. The steep peak in the beginning shows the high concentration of low expression values in most genes




