---
title: "Task3 - Exploratory Data Analysis"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
library(here)
sample <- read.csv2(here("data", "sample.tsv"), sep="\t", header=TRUE)


```
## Data Overview 
```{r}
head(sample)
dim(sample)
sum(sample$count==0)
```
```{r}
min(sample$count)
max(sample$count)
mean(sample$count)
sd(sample$count)
```
# Plotting
```{r}
plot(sample$count)
barplot(sample$count)
stripchart(sample$count)
boxplot(sample$count)
hist(sample$count)
plot(density(sample$count))
```

### Methodology and Code Implementation
* Row 8: `read.csv2`imports the data of the chosen file (sample.tsv); the parameters specify that the file uses tabs (sep="\t") as separators and contains column names in the first row (header = TRUE)   
* Row 13: `head()`shows the first six rows of the dataset to provide a quick look into the data structure   
* Row 14: `dim()` shows the dimensions of the dataset (number of rows and columns)   
* Row 15: `sum(sample$count==0)`calculates the total number of genes without expression    
* Row 20 & 21: `min()`/ `max()` identify the lowest/ highest expression values in the column "count"   
* Row 22: `mean()` calculates the average expression level   
* Row 23: `sd()` calculates the standard deviation (shows dispersion around the average)   
* Row 27-32: create 6 different plot as visual representation of the values

### Statistical Analysis
* Data Preview: shows the first 6 genes and its amount of expression. For example, in these 6 genes, the highest expression is in gene ICR1 with 194 counts.   
* General statistics: Result of `dim()` with **7,126** rows (= genes) and 2 columns (gene_id and counts)
          **1,090** genes have an expression value of zero 
          the counts range from a min of **0** counts to a max. **194,695** counts (no transciption under these circumstances)
          the mean is approximately **749** counts, the sd **3,837** counts(high sd indicates a high level of dispersion)   
          
### Visual Interpretation          
* Indexplot: each point represents one gene. The x-axis could be the row number in the table, the y-axis the expression count
Most genes have low expression (**under 30,000** counts) while few outliers are expressed highly (**>100,000** counts)  
* Barplot: shows a expression as a bar for each gene in the dataset. its shows the same like the Indexplot, just a different type of graphics/ visualization.  
* Stripchart: is a one-dimensional scatter plot, that stacks the expression values on an horizontal axis. shows high density and gaps in the expression of genes of this dataset.   
* Boxplot: the box (not visible) represents the middle 50% of the data. looks like a line in the bottom part, because the mean and most values have similar expression rates   
* Histogram: gathers frequencies of different expression rates in columns. Each column represents 10,000 counts in expression rate. the first column contains almost all values   
* Density plot: is a smoothed version of the Histogram showing the probability distribution. The steep peak in the beginning shows the high concentration of low expression values in most genes





